---
title: "Acetaminophen-PBPK model"
author: "Nan-Hung Hsieh"
date: "`r Sys.Date()`"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acetaminophen-PBPK model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(pksensi)
mcsim_install()
#library(kableExtra)
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.height=5, 
  fig.width=8,
  comment = "#>"
)
```

The aim of this section is to reproduce our previous published [@fphar201800588] result of global sensitivity analysis for acetaminophen PBPK model through `pksensi`. The model code are included in this package and can be generated through `pbpk_apap_model()`. We applied global sensitivity analysis to the original published model with 21 model parameters [@s13318-015-0253-x]. The descriptions of each parameter and the sampling ranges are list in Table 1.

```{r, echo=F, eval=T}
#Nominal value
Tg <- log(0.23)
Tp <- log(0.033)
CYP_Km <- log(130)
SULT_Km_apap <- log(300)
SULT_Ki <- log(526)
SULT_Km_paps <- log(0.5)
UGT_Km <- log(6.0e3)
UGT_Ki <- log(5.8e4)
UGT_Km_GA <-log(0.5)
Km_AG <- log(1.99e4)
Km_AS <- log(2.29e4)

r <- 2 # exp(2)/exp(-2) ~ 54

x <- c("Tg", "Tp", "CYP_Km", "CYP_VmaxC",
       "SULT_Km_apap","SULT_Ki","SULT_Km_paps","SULT_VmaxC",
       "UGT_Km","UGT_Ki","UGT_Km_GA","UGT_VmaxC",
       "Km_AG","Vmax_AG","Km_AS","Vmax_AS",
       "kGA_syn","PAPS_syn", "CLC_APAP","CLC_AG","CLC_AS")
y <- c("Gatric emptying time constant",
       "GI perfusion time constant",
       "Cytochrome P450 metabolism, Km",
       "Cytochrome P450 metabolism, VMax",
       "Sulfation pathway acetaminophen, Km",
       "Sulfation pathway substrate inhibition, Ki",
       "Sulfation pathway PAPS, Km",
       "Sulfation pathway acetaminophen, Vmax",
       "Glucronidation pathway acetaminophen, Km",
       "Glucronidation pathway substrate inhibition, Ki",
       "Glucronidation pathway GA, Km",
       "Glucronidation pathway acetaminophen, Vmax",
       "APAP-G hepatic transporter, Km",
       "APAP-G hepatic transporter, Vmax",
       "APAP-S hepatic transporter, Km",
       "APAP-S hepatic transporter, Vmax",
       "UDPGA synthesis",
       "PAPS synthesis",
       "APAP clearance",
       "APAP-G clearance",
       "APAP-S clearance")
z <- c("$h$", "$h$", "$\\mu{M}$", "$\\mu{mole}/h\\cdot{BW}^{0.75}$",
       "$\\mu{M}$", "$\\mu{M}$", "$-$", "$\\mu{mole}/h\\cdot{BW}^{0.75}$",
       "$\\mu{M}$", "$\\mu{M}$", "$-$", "$\\mu{mole}/h\\cdot{BW}^{0.75}$",
       "$\\mu{M}$", "$\\mu{mole}/h$", "$\\mu{M}$", "$\\mu{mole}/h$",
       "$1/h$", "$1/h$",
       "$L/h\\cdot{BW}^{0.75}$", "$L/h\\cdot{BW}^{0.75}$", "$L/h\\cdot{BW}^{0.75}$")
min <- c(round(Tg-r, 3), round(Tp-r, 3), round(CYP_Km-r), round(log(0.14), 3),
         round(SULT_Km_apap-r, 3), round(SULT_Ki-r, 3), round(SULT_Km_paps-r), log(1),
         round(UGT_Km-r, 3), round(UGT_Ki-r, 3), round(UGT_Km_GA-r), log(1),
         round(Km_AG-r, 3), round(log(1.09e3), 3), round(Km_AS-r), round(log(1.09e3), 3),
         log(1), log(1), round(log(2.48e-3), 3), round(log(2.48e-3), 3), round(log(2.48e-3), 3))
max <- c(round(Tg+r, 3), round(Tp+r, 3), round(CYP_Km+r), round(log(2900), 3),
         round(SULT_Km_apap+r, 3), round(SULT_Ki+r, 3), round(SULT_Km_paps+r), round(log(22026), 3),
         round(UGT_Km+r, 3), round(UGT_Ki+r, 3), round(UGT_Km_GA+r), round(log(22026), 3),
         round(Km_AG+r, 3), round(log(3.26e6), 3), round(Km_AS+r), round(log(3.26e6), 3),
         round(log(4.43e5), 3), round(log(4.43e5),3), 
         round(log(2.718), 3), round(log(2.718), 3), round(log(2.718), 3))

df <- data.frame(x, y, z, min, max)
names(df) <- c("Parameter","Description", "Unit", "Min", "Max")

#if (require(kableExtra)) {
  knitr::kable(df, format = 'html', align=c("l","l","l", "c", "c"), 
               caption = "Table 1 Description of sampling range of model parameter") 
  #%>% kableExtra::add_footnote(c("The parameter valur are showed in log-transformed scale."), notation = "number")    
#}

```

Same as the example of one-compartment PBTK model. The model parameters and the corresponding sampling range should be defined to create the parameter matrix. Previously, the probability distribution of model parameter were set to either truncated normal or uniform distribution when the parameter have informative prior information or not. To rapidly reach the acceptance convergence, we apply uniform distribution for all testing parameters. The range of informative parameter are setting to 4-times log-scaled (approximate 54.6 times difference between minimum and maximum). The nomial value of informative model parameters were defin as:

```{r}
# Nominal value
Tg <- log(0.23)
Tp <- log(0.033)
CYP_Km <- log(130)
SULT_Km_apap <- log(300)
SULT_Ki <- log(526)
SULT_Km_paps <- log(0.5)
UGT_Km <- log(6.0e3)
UGT_Ki <- log(5.8e4)
UGT_Km_GA <-log(0.5)
Km_AG <- log(1.99e4)
Km_AS <- log(2.29e4)

rng <- 2 
```

The wide range of parameter value might cause the computational error in solver. One of effective solution is to adjust the value of relative and absoulte error tolerance to control the error appearance by resetting these parameter in lower value. The `generate_infile()` provide the arguments of `rtol` and `atol` that can be adjusted to  . However, the modification will slow down the computational speed. Therefore, the alternative method to prevent the computational error is to detect the crucial parameter range that cause the problem.  

We adjust the range of `SULT_VmaxC` and `UGT_VmaxC` from U(0, 15) to U(0, 10). The relative and absoulte error tolerance are set to 1e-7 and 1e-9, respectively, to prevent the computational error in MCSim in this example,

```{r}
params <- c("lnTg", "lnTp", "lnCYP_Km","lnCYP_VmaxC",
           "lnSULT_Km_apap","lnSULT_Ki","lnSULT_Km_paps","lnSULT_VmaxC",
           "lnUGT_Km","lnUGT_Ki","lnUGT_Km_GA","lnUGT_VmaxC",
           "lnKm_AG","lnVmax_AG","lnKm_AS","lnVmax_AS",
           "lnkGA_syn","lnkPAPS_syn", "lnCLC_APAP","lnCLC_AG","lnCLC_AS")
q <- "qunif"
q.arg <-list(list(Tg-rng, Tg+rng),
             list(Tp-rng, Tp+rng),
             list(CYP_Km-rng, CYP_Km+rng),
             list(-2., 5.),
             list(SULT_Km_apap-rng, SULT_Km_apap+rng),
             list(SULT_Ki-rng, SULT_Ki+rng),
             list(SULT_Km_paps-rng, SULT_Km_paps+rng),
             list(0, 10), #U(0.15)
             list(UGT_Km-rng, UGT_Km+rng),
             list(UGT_Ki-rng, UGT_Ki+rng),
             list(UGT_Km_GA-rng, UGT_Km_GA+rng),
             list(0, 10), #U(0.15)
             list(Km_AG-rng, Km_AG+rng),
             list(7., 15),
             list(Km_AS-rng, Km_AS+rng),
             list(7., 15),
             list(0., 13),
             list(0., 13),
             list(-6., 1),
             list(-6., 1),
             list(-6., 1))

times <- seq(from = 0.1, to = 12.1, by = 0.2)
set.seed(1234)
x <- rfast99(params = params, n = 512, q = q, q.arg = q.arg, replicate = 5) 
```

Same as the initial step in one-compartment PBTK model.
Then, use `compile_model()` to generate the executable files (`mcsim.pbpk_apap`). 

```{r}
mName <- "pbpk_apap"
pbpk_apap_model()
compile_model(mName, application = "mcsim")
```

To improve the computational speed, this case only use MCSim to solve

```{r}
vars <- c("lnCPL_APAP_mcgL", "lnCPL_AG_mcgL", "lnCPL_AS_mcgL")
conditions <- c("mgkg_flag = 0",
                "OralExp_APAP = NDoses(2, 1, 0, 0, 0.75)",
                "OralDur_APAP = 0.75",
                "OralDose_APAP_mg = 1000.0",
                "IVExp_APAP = 0.",
                "IVDose_APAP_mg = 0.")
generate_infile(params = params,
                vars = vars,
                time = times, 
                condition = conditions, 
                rtol = 1e-6) 
y <- solve_mcsim(x, mName = mName, 
                 params = params,
                 vars = vars,
                 time = times,
                 condition = conditions,
                 generate.infile = F)
tell2(x,y)
```

The plotting function can output the result of time-dependent sensitivity measurement. In this case, the target variable are the concentration of APAP, AG and AS in plasma.  

```{r, fig.height=8, fig.width=8, fig.cap = 'Figure 2. '}
plot(x, vars = "lnCPL_AG_mcgL")
```


```{r}
check(x, SI.cutoff = 0.05)
```


```{r}
heat_check(x, order = "interaction")
```

```{r}
heat_check(x, order = "total order")
```

```{r}
heat_check(x, index = "CI")
```



## References
